Forest canopies are better generators of biodiversity in the tropics

### 2025-07-14 manuscript.Rmd
ms conclusion

---
date: "`r Sys.Date()`"
author: "Your Name"
title: "Forest canopies are better generators of biodiversity in the tropics"
output: 
  officedown::rdocx_document:
    mapstyles:
      Normal: ['First Paragraph']
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.cap = TRUE)
library(officedown)
library(officer)

@@

ft <- fp_text(shading.color='#EFEFEF', bold = TRUE)
```

```{r}
# analysis scripts
# storing all figures in figures/ms_figures/
# other figures are from older versions

"scripts/03_analyses/01_canopyheight_richness_relationships.R" # canopy height richness models and plots
```

@@

For centuries, ecologists have been intrigued by the role of vertical landscape features in biogeographic patterns from local to global scales [@vonhumboltEssayGeographyPlants2013; @macarthurBirdSpeciesDiversity1961; @bashamVerticalStratificationPatterns2023a; @xingEcologicalPatternsProcesses2023a]. In 1799, Alexander von Humbolt began an expedition that would ultimately canonize the role of mountains in global richness patterns [@vonhumboltEssayGeographyPlants2013; @HumboldtsLegacy2019]. [Yet until recently, hypotheses failed to explain the extraordinary richness of mountains particularly in the tropics.] The vertical nature of mountains provides more land surface per unit area than flat regions and produces steep climate gradients over short distances that allow rapid rates of species turnover. Combined, these factors allow mountains to generate an extraordinarily high biodiversity for a given area relative to flat and climatically homogeneous lowlands [@rahbekHumboldtsEnigmaWhat2019]. Yet mountains are better generators of biodiversity in the tropics because high latitude mountains have lower elevation snow lines that truncate the area of land suitable for diverse communities, and greater seasonal temperature variation, which reduces the climatic distinctness of elevational bands relative to tropical mountains [@rahbekHumboldtsEnigmaWhat2019; @vonhumboltEssayGeographyPlants2013].

NEED TO MENTION HUMBOLT'S ENIGMA SOMEWHERE IN FIRST PARAGRAPH AND THAT THE MAIN EXPLANATIONS ARE TOPOGRAPHY AND CLIMATE

Over a century later, Robert MacArthur highlighted the role of another vertical dimension in promoting high biodiversity: the vertical structure of vegetation, which is increasingly being recognized as critical to our understanding of biogegoraphic patterns [@macarthurBirdSpeciesDiversity1961]. Tall, multi-layered forests generate biodiversity through similar mechanisms to mountains. By increasing habitat volume and producing vertical zones of microclimate and resource availability, tall forests allow species to vertically stratify habitat akin to multi-floored apartments [@bashamVerticalStratificationPatterns2023a; @xingEcologicalPatternsProcesses2023a; @macarthurLimitingSimilarityConvergence1967; @macarthurBirdSpeciesDiversity1961; @larueDiversityVolumeRelationships2023]. Thus, taller forests increase niche opportunities, which may allow additional species to join the assemblage and lead to greater richness within a given planimetric area compared to regions with shorter vertical vegetation structure [@xingEcologicalPatternsProcesses2023a; @macarthurBirdSpeciesDiversity1961; @macarthurLimitingSimilarityConvergence1967]. This vertical stacking of species has been described as a mechanistic link between canopy height and species richness, and is thought to contribute to exceptionally high biodiversity in the tropics where canopies are often tall and structurally complex [@oliveiraVerticalStratificationInfluences2019; @rollLinkingVertebrateSpecies2015; @fengForestCanopyHeight2020]. Yet positive impacts of canopy height on richness are not consistent globally, with often weaker relationships in high latitude regions where forests of comparable vertical profiles to those in the tropics harbor fewer species [@larueDiversityVolumeRelationships2023; @macarthurBirdSpeciesDiversity1961; @rollLinkingVertebrateSpecies2015; @fengForestCanopyHeight2020]. - a vertical dimension to Humbolt's enigma

Despite decades of ecological inquiry into the vertical dimensions of habitat structure and biodiversity [@macarthurBirdSpeciesDiversity1961; @scheffersIncreasingArborealityAltitude2013; @xingEcologicalPatternsProcesses2023a], the mechanisms that help or hinder the capacity for tall forests to generate biodiversity remain poorly understood yet are critically important for understanding biogeographic patterns, impacts of climate and landuse change on biodiversity, and the conservation value of forests. Following upon explanations to Humbolt's enigma [@rahbekHumboldtsEnigmaWhat2019], we fill this gap by examining how vegetative complexity and climate may play key roles in mediating the relationship between canopy height and species richness by impacting the capacity for tall canopies to support vertically stratified and species rich arboreal assemblages. For example, the denser layering of vegetation in tropical forests may provide more niche opportunities, allowing tall canopies to support more biodiversity than canopies of equivalent height in temperate regions [CITE or CONCEPT FIG]. Moreover, densely layered vegetation produces vertical microclimate gradients via interactions with macroclimate conditions, solar radiation, and wind speed that are characterized by humid and thermally stable understories to dry and thermally variable canopies that experiences both hot and cold temperature extremes. Physiological constraints imposed on arboreal life strategies by these more extreme conditions, combined with the distinctness of vertical climate bands may impact species turnover from the ground to the canopy and thus the capacity for taller canopies to support increasing richness [@bashamVerticalStratificationCollapses2020] [CITE]. Yet only recently have we been able to address these hypotheses, as our ability to measure vegetation structure and quantify its use by animals globally [AND MODEL MICROCLIMATES GLOBALLY] increased with the collation of large ecological databases and the launch of satellite missions to measure forest height and structural complexity [@popkinSpaceLaserMap2018; @mouraPhylogenyinformedCharacterisationGlobal2024a].

In this study we combine range maps of nearly 24,000 terrestrial vertebrate species and comprehensive databases of vertical habitat use, global macroclimate and forest structure data, and process-based microclimate models to examine how forest structure and climate may mediate the capacity for tall forests to support species rich assemblages via impacts on the vertical structure of vertebrate communities. We first investigated the capacity for taller forests to support increasing species richness in forests and less densely vegetated wooded biomes across latitudes. If the vertical structure of species assemblages mediates the relationship between canopy height and species richness, then we expect positive correlations between the verticality of assemblages (i.e., their average vertical niche position) and species richness. [AND THAT CANOPY HEIGHT WILL EXPLAIN LITTLE VARIATION IN RICHNESS ONCE VERTICALITY HAS BEEN ACCOUNTED FOR]

and expected that tropical canopies would exhibit the fastest rates of increase in richness with canopy height. We then examined the mechanistic processes driving this pattern.  EXPLANATION OF TETRAPODTRAITS AND ELTONTRAITS

, including TetrapodTraits and EltonTraits, to map species richness and the vertical structure of species assemblages in wooded biomes across the globe.

To fill this gap, we combine range maps of nearly 24,000 terrestrial vertebrate species with the TetrapodTraits and EltonTraits databases to map species richness and the vertical structure of species assemblages in wooded biomes across the globe. We first investigated the capacity for taller forests to support greater species richness in forests and less densely vegetated wooded biomes across latitudes and expected that tropical canopies would exhibit the fastest rates of increase in richness with canopy height. We then examined the mechanistic processes driving this pattern. If vertical stratification contributes to the capacity for tall forests to support diverse assemblages, we expect that species rich assemblages will exhibit high verticality (i.e., a high average vertical niche position) and that environmental factors limiting verticality will limit species richness. To test this hypothesis, we examined correlations between assemblage verticality and species richness. We then modeled assemblage verticality of the four vertebrate classes by macroclimate and vegetative structure and investigated the microclimatic processes that underlie these macroclimatic patterns. While three-dimensional climate data is not currently available globally given the large computational requirements of resolving micrometeorological processes within forest canopies, we modeled vertical microclimate profiles in six sites across the globe within tropical, temperate, and boreal forests by integrating forest structure data derived from NASA's Global Ecosystem Dynamics Investigation (GEDI) mission with high temporal resolution macroclimate data in a process-based microclimate model.

@@

To evaluate the capacity for tall canopies to generate biodiversity, we modeled vertebrate richness as a function of canopy height for all vertebrates and each vertebrate class globally. We then mapped the residuals of these models to determine where canopy height over- or under-predicted species richness. We then produced separate models for tropical, temperate, and boreal regions in forests and more sparsely vegetated habitats, including woodlands, savannas, and shrublands, to examine variation in the capacity for wooded habitats of the same height to generate biodiversity and variation in the rate of increasing richness with a given increase in canopy height across different regions.

Globally, taller trees, which increase habitat volume, support more species (Fig. S\@ref(fig:global-chrich)). However, global models of vertebrate richness by canopy height overpredict richness at high latitudes (negative residuals) and underpredict richness in the tropics (positive residuals) (Fig. \@ref(fig:fig1), S\@ref(fig:resids-taxa)). Therefore, canopies of equivalent height generally harbor more species in the tropics than in temperate and boreal regions. Furthermore, increasing canopy height yields greater gains in richness in the tropics than at higher latitudes (Fig. \ref(fig:fig1)). For example, an increase in canopy height from 15 m to 25 m increases richness by `r round(ratedifs$dif[ratedifs$biome2 == "Tropical forest" & ratedifs$taxa=="Birds"])` birds and `r round(ratedifs$dif[ratedifs$biome2 == "Tropical forest" & ratedifs$taxa=="Amphibians"])` amphibians in tropical forests and by only `r round(ratedifs$dif[ratedifs$biome2 == "Boreal" & ratedifs$taxa=="Birds"])` birds and `r round(ratedifs$dif[ratedifs$biome2 == "Boreal" & ratedifs$taxa=="Amphibians"])` amphibians in boreal regions.  Within the tropics, increasing canopy height generally yields greater gains in richness in forests than in drier woodlands, savannas, and shrublands, though baseline richness in areas with short vegetation are similar (Fig. \@ref(fig:fig1)). For the same increase in canopy height, the increase in bird richness in tropical forests exceeds the increase in bird richness in tropical woodlands, savannas, and shrublands by `r round(ratedifs$dif[ratedifs$biome2 == "Tropical forest" & ratedifs$taxa=="Birds"]) - round(ratedifs$dif[ratedifs$biome2 == "Tropical woodland, savanna, and shrubland" & ratedifs$taxa=="Birds"])` species. The difference in the capacity for tall canopies to generate biodiversity between densely and sparsely vegetated tropical regions is particularly evident in Madagascar where the global model underpredicts vertebrate richness in the humid forests of the east coast and overpredicts richness in the drier woodlands and shrublands of the southwest (Fig. \@ref(fig:fig1)).

Minor deviations from these patterns may indicate variations in evolutionary drivers of species richness patterns.
  [- This pattern is particularly stark in the western hemisphere. In the eastern hemisphere, the pattern is weaker where evolutionary forces may drive overprediction of richness in parts of the lowland African tropics and Indo-Malaysian. In Indo-Malaysia, island size may reduce vertebrate richness relative to expectations by canopy height.]

However, the described patterns are broadly consistent among taxa and mirror findings that for a given area, mountains have higher richness in the tropics [@rahbekHumboldtsEnigmaWhat2019]. Thus, increases in habitat volume via either mountains or forests have greater impacts on richness in the tropics. In other words, taller canopies generate biodiversity globally, but are more effective generators of biodiversity in the tropics.

# The mediating effect of verticality
The vertical layering of species within forests may mediate the capacity for tall forests to support rich assemblages. As forests increase in height, additional arboreal species may be added to the assemblage in novel canopy space [@macarthurBirdSpeciesDiversity1961; @rollLinkingVertebrateSpecies2015; @larueDiversityVolumeRelationships2023]. An increase in the number of arboreal species without an equivalent increase in terrestrial or fossorial species increases the verticality of the assemblage, which represents the skew in the vertical distribution of species toward the canopy [@oliveiraVerticalStratificationInfluences2019; @scheffersVerticalArborealityHorizontal2017]. Within wooded ecosystems, we therefore expect regions exhibiting high verticality to support species rich assemblages and steep increases in richness with canopy height [@oliveiraVerticalStratificationInfluences2019]. To quantify the verticality of an assemblage, we use the aforementioned trait databases, assigning verticality scores to each species based on their occurrence in fossorial (0), terrestrial (0.5), and arboreal (1) habitats [@oliveiraVerticalStratificationInfluences2019; @scheffersVerticalArborealityHorizontal2017]. For species occurring in multiple habitats, we averaged scores across habitats. We then calculated verticality as the standard effect size of the average verticality score across species in an assemblage, using 100 simulations to determine if assemblages are more vertical than expected by random chance given their richness and the evolutionary history of the biogeographic realm [@oliveiraVerticalStratificationInfluences2019] (Supplementary methods).

Verticality peaked in the tropics and declined toward the poles, exhibiting strong correlations with verticality across taxa (Fig. \@ref(fig:fig2)). Moist tropical forests contained assemblages comprised of x% to x% species exhibiting arboreal strategies (SUPP FIG), corresponding to some of the most species rich assemblages. Though temperate and boreal regions contain tall forests, albeit not quite reaching the heights of the tallest tropical canopies, they contain assemblages with much lower verticality, and higher proportions of species exhibiting terrestrial and fossorial strategies. This pattern suggests that the capacity for taller canopies to support higher richness declines when canopies cannot support highly vertical species assemblages. Understanding environmental factors that constrain the evolution of diverse arboreal assemblages is therefore key to unraveling the vertical dimension to Humbolt's enigma.

# Vegetation structure and canopy climates impact the capacity for tall canopies to support vertical communities
To examine the environmental factors influencing assemblage verticality, we used spatial linear mixed models to model assemblage verticality for each vertebrate class as a function of canopy height and its interaction with macroclimatic factors that may constrain or support arboreal life strategies, including minimum temperature of the coldest month, maximum temperature of the warmest month, precipitation of the driest month, and precipitation of the warmest month, as well as with macroclimatic factors that may influence the distinctness of vertical climate banding, including diurnal temperature range (Supplementary methods). We additionally included vegetation density in the models, which impacts the availability of habitat volume. While foliage height diversity and temperature seasonality also impact niche opportunities and the distinctness of climate banding, these variables where strongly correlated with canopy height and minimum temperatures, respectively, and were therefore excluded from the models. The macroclimate variables included in these models represent climate conditions experienced in open habitats rather than the microclimate conditions across forest profiles [@defrenneWeatherStationsLack2016; @bramerAdvancesMonitoringModelling2018]. While using macroclimate data was necessary due to the lack of global vertically resolved microclimate data, its use inhibits our ability to understand how microclimates experienced by forest-dwelling organisms impact community structure [@klingesProximalMicroclimateMoving2024; @zellwegerForestMicroclimateDynamics2020]. We therefore additionally modeled vertical microclimate profiles in six forests from tropical to boreal latitudes by integrating information on the vertical distribution of vegetation derived from NASA's GEDI spaceborne lidar mission [@dubayahGlobalEcosystemDynamics2020] with ERA5 macroclimate data at a high temporal resolution [@hersbachERA5HourlyData2023] in a process-based microclimate model that resolves within-canopy processes to predict vertical climate gradients [@macleanMicroclimfFastCanopy2024]. To reduce variability in vertical profiles not attributed to latitudinal differences, we restricted microclimate models to points below 500 m a.s.l and that represented the top 75th to 98th quantiles of canopy height within a 1° x 1° grid cell. By estimating the climate conditions experienced by forest-dwelling organisms, these models provide insight into the microclimatic processes that underlie macroclimate patterns and allow canopies to support vertically stratified and species rich assemblages.

Our models of verticality indicate that taller canopies support highly vertical vertebrate assemblages, while dry, and thermally variable macroclimate conditions, including extreme hot and cold temperatures limit verticality and mediate the capacity for tall forests to support vertical assemblages, albeit to varying degrees across vertebrate classes (Fig. \@ref(fig:fig3)). These patterns arise due to impacts of spatial climate variation on vegetation structure and physiological constraints to arboreal lifestyles as well as impacts of temporal climate variation on the distinctness of vertical climate bands.

## Climate produces structural constraints to verticality
Climate impacts the structural complexity of vegetation, which plays a key role in the the availability of arboreal niche opportunities and is typically associated with high animal richness [@xingEcologicalPatternsProcesses2023a; @daviesAdvancesAnimalEcology2014]. High dry season precipitation and thus low precipitation seasonality may support highly vertical assemblages by enabling development of tall and structurally complex forests with densely layered vegetation and generally high tree diversity [@ehbrechtGlobalPatternsClimatic2021a; @decontoCharacterizingStructuralComplexity2024; @hakkenbergClimateMediatesRelationship2021]. Wet climates also support diverse epiphyte communities, which provide additional microhabitats in the canopy that contribute to high arboreal richness [@taylorVascularEpiphytesContribute2022; @thomsenSecondaryFoundationSpecies2018; @ellwoodDoublingEstimateInvertebrate2004]. Hotspots of structural complexity that coincide with some of the tallest forests, including the Choco-Darien, the Amazon, Borneo, and New Guinea [@ehbrechtGlobalPatternsClimatic2021a; @decontoCharacterizingStructuralComplexity2024], thus provide opportunities for fine-scale niche partitioning, yielding species rich assemblages [@camargoForestVerticalComplexity2018; @gouveiaForestStructureDrives2014; @oliveiraVerticalStratificationInfluences2019]. In contrast, lower structural complexity in parts the Afrotropics [@ehbrechtGlobalPatternsClimatic2021a] may reduce niche opportunities for arboreal species, yielding less vertical assemblages and lower richness than expected based on the height of the canopy (Fig. \@ref(fig:fig1)).

>>>>

---
bibliography: "`r rbbt::bbt_write_bib('r-refs.bib', overwrite = TRUE)`"
csl: science.csl
output: 
  officedown::rdocx_document:
    reference_docx: "style-guide.docx"
    mapstyles:
      Normal: ['First Paragraph']
plots:
  caption:
    style: Image Caption
    pre: 'Figure '
    sep: ': '
header-includes:
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \rhead{Global Verticality}
- \begin{document}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.cap = TRUE, warning=FALSE, message=FALSE, error=FALSE)
library(officedown)
library(officer)
library(tidyverse)
library(spatialreg)
library(data.table)
library(flextable)
library(knitr)
library(chunkhooks)

@@

ft <- fp_text(shading.color='#EFEFEF', bold = TRUE)

hook_figure_unit("mm")
```

```{r}
# analysis scripts
# storing all figures in figures/ms_figures/
# other figures are from older versions
```

@@

**Forest canopies are better generators of biodiversity in the tropics**

Lydia G. Soifer^1\*^, David P. Edwards^2,3^, Brunno F Oliveira^4,5^, Mario R Moura^6^, Brett R. Scheffers^4^

1 School of Natural Resource and Environment, University of Florida, USA

2 Department of Plant Sciences and Centre for Global Wood Security, University of Cambridge, Cambridge, UK

3 Conservation Research Institue, University of Cambridge, Cambridge, UK

4 Centre de Synthèse et d’Analyse sur la Biodiversité, Montpellier, France

5 Department of Wildlife Ecology and Conservation, University of Florida, USA

6 Departamento de Sistemática e Ecologia, Universidade Federal da Paraíba, João Pessoa, Brazil

Email addresses:

^\*^ lysoifer@gmail.com (corresponding author)

dpe29@cam.ac.uk

brunno.oliveira@me.com

mariormoura@gmail.com

brett.scheffers@ufl.edu

# Abstract

For centuries, ecologists have been intrigued by the role of vertical landscape features in biogeographic patterns from local to global scales [@vonhumboltEssayGeographyPlants2013; @macarthurBirdSpeciesDiversity1961; @bashamVerticalStratificationPatterns2023a; @xingEcologicalPatternsProcesses2023a]. In 1799, Alexander von Humbolt began an expedition that would ultimately canonize the role of mountains in global biodiversity patterns as generators of species richness, particularly in tropical regions [@vonhumboltEssayGeographyPlants2013; @HumboldtsLegacy2019; @rahbekHumboldtsEnigmaWhat2019]. The extraordinarily high biodiversity of mountains and their enhanced capacity to generate species richness in the tropics has been termed "Humbolt's enigma", and can be attributed to topographic and climatic variation [@rahbekHumboldtsEnigmaWhat2019]. The vertical nature of mountains provides more land surface per unit area than flat regions and produces steep climate gradients over short distances that allow rapid rates of species turnover. Yet mountains are better generators of biodiversity in the tropics because high latitude mountains have lower elevation snow lines that truncate the area of land suitable for diverse communities, and greater seasonal temperature variation, which reduces the climatic distinctness of elevational bands relative to tropical mountains [@rahbekHumboldtsEnigmaWhat2019; @vonhumboltEssayGeographyPlants2013; @janzenWhyMountainPasses1967].

Over a century later, Robert MacArthur highlighted the role of another vertical dimension in promoting high biodiversity: the vertical structure of vegetation, which is increasingly being recognized as critical to our understanding of biogeographic patterns [@macarthurBirdSpeciesDiversity1961; @fredstonReimaginingSpeciesMove2025; @xingEcologicalPatternsProcesses2023a]. Tall, multi-layered forests generate biodiversity through similar mechanisms to mountains. By increasing habitat volume and producing vertical zones of microclimate and resource availability, tall forests allow species to vertically stratify habitat akin to multi-floored apartments [@bashamVerticalStratificationPatterns2023a; @xingEcologicalPatternsProcesses2023a; @macarthurLimitingSimilarityConvergence1967; @macarthurBirdSpeciesDiversity1961; @larueDiversityVolumeRelationships2023]. Thus, taller forests increase niche opportunities, which may allow additional species to join the assemblage and lead to greater richness within a given planimetric area compared to regions with shorter vertical vegetation structure [@xingEcologicalPatternsProcesses2023a; @macarthurBirdSpeciesDiversity1961; @macarthurLimitingSimilarityConvergence1967]. This vertical stacking of species has been described as a mechanistic link between canopy height and species richness, and is thought to contribute to exceptionally high biodiversity in the tropics where canopies are often tall and structurally complex [@oliveiraVerticalStratificationInfluences2019; @rollLinkingVertebrateSpecies2015; @fengForestCanopyHeight2020; @decontoCharacterizingStructuralComplexity2024]. Yet positive impacts of canopy height on richness are not consistent globally, with often weaker relationships in high latitude regions where forests of comparable vertical profiles to those in the tropics harbor fewer species [@larueDiversityVolumeRelationships2023; @macarthurBirdSpeciesDiversity1961; @rollLinkingVertebrateSpecies2015; @fengForestCanopyHeight2020]. The capacity for forests to generate biodiversity via vertical stratification, yet the variation in their capacity to do so globally therefore presents a vertical dimension to Humbolt's enigma.

Despite decades of ecological inquiry into the vertical dimensions of habitat structure and biodiversity [@macarthurBirdSpeciesDiversity1961; @scheffersIncreasingArborealityAltitude2013; @xingEcologicalPatternsProcesses2023a; @rollLinkingVertebrateSpecies2015], the mechanisms that help or hinder the capacity for tall forests to generate biodiversity remain poorly understood yet are critically important for understanding biogeographic patterns, impacts of climate and landuse change on biodiversity, and the conservation value of forests. Following upon explanations to Humbolt's enigma [@rahbekHumboldtsEnigmaWhat2019], we fill this gap by examining how vegetative complexity and climate may play key roles in mediating the relationship between canopy height and species richness by impacting the capacity for tall canopies to support vertically stratified and species rich arboreal assemblages. For example, denser layering of vegetation in tropical forests may provide more niche opportunities, allowing tall canopies to support more biodiversity than canopies of equivalent height in temperate regions [@daviesAdvancesAnimalEcology2014; @oliveiraVerticalStratificationInfluences2019]. Moreover, densely layered vegetation produces vertical microclimate gradients via interactions with macroclimate conditions, solar radiation, and wind speed that are characterized by humid and thermally stable understories to dry and thermally variable canopies that experiences both hot and cold temperature extremes. Physiological constraints imposed on arboreal life strategies by these more extreme conditions, combined with the distinctness of vertical climate bands may impact species turnover from the ground to the canopy and thus the capacity for taller canopies to support increasing richness [@bashamVerticalStratificationCollapses2020; @oliveiraVerticalStratificationInfluences2019]. Yet only recently have we been able to address these hypotheses, as our ability to quantify variation in vegetation structure and microclimates globally and examine their impacts on biogeographic patterns in three-dimensional space has increased with the launch of satellite missions to measure forest height and structural complexity [@popkinSpaceLaserMap2018], the development of microclimate models that can estimate 3D microclimates any where on earth [@macleanMicroclimfFastCanopy2024], and the collation of large ecological databases on vertical habitat use [@mouraPhylogenyinformedCharacterisationGlobal2024a; @wilmanEltonTraitsSpecieslevelForaging2014].

In this study we combine range maps of over 24,000 terrestrial vertebrate species with comprehensive data on vertical habitat use from the TetrapodTraits and EltonTraits databases [@mouraPhylogenyinformedCharacterisationGlobal2024a; @wilmanEltonTraitsSpecieslevelForaging2014] to examine how forest structure and climate may mediate the capacity for tall forests to support species rich assemblages via impacts on the vertical structure of vertebrate communities. We first investigated the capacity for taller canopies to support increasing species richness in forests and less densely vegetated woody biomes across latitudes. We find that tall canopies are better generators of biodiversity in the tropics, particularly in moist forests (Fig. \ref(fig:fig1)). We then examined the processes driving this pattern using a combination of global models of vertical assemblage structure parameterized with macroclimate and forest structure data and process-based microclimate models [@macleanMicroclimfFastCanopy2024; @macleanMicropointCanopyPoint2025] for smaller regions that integrate forest structure data derived from NASA's Global Ecosystem Dynamics Investigation (GEDI) [@dubayahGlobalEcosystemDynamics2020] with high temporal resolution macroclimate data [@hersbachERA5HourlyData2023]. While the large computational requirements of resolving micrometeorological processes within forest canopies inhibit the use of global 3D climate data in our analysis, site-specific models spanning six regions across tropical, temperate, and boreal forests provide insight into the fine-scale processes underlying large-scale patterns.

@@

To evaluate the capacity for tall canopies to generate biodiversity, we fit negative binomial regression models of vertebrate richness as a function of canopy height globally for all vertebrate combined and each vertebrate class (birds, mammals, reptiles, and amphibians). We then mapped the deviance residuals of these models to determine where canopy height over- or under-predicted species richness. To examine variation in the capacity for tall canopies to support rich assemblages across latitudes, we produces separate models for tropical, temperate, and boreal regions in forests and more sparsely vegetated habitats, including woodlands, savannas, and shrublands.

Globally, taller trees, which increase habitat volume, support more species (Fig. S\@ref(fig:global-chrich)). However, global models of vertebrate richness by canopy height overpredict richness at high latitudes (negative residuals) and underpredict richness in the tropics (positive residuals) (Figs. \@ref(fig:fig1)c, S\@ref(fig:resids-taxa)). Therefore, canopies of equivalent height generally harbor more species in the tropics than in temperate and boreal regions. Furthermore, increasing canopy height yields greater gains in richness in the tropics than at higher latitudes (Fig. \ref(fig:fig1)d). For example, an increase in canopy height from 15 m to 25 m increases richness by `r round(ratedifs$dif[ratedifs$biome2 == "Tropical forest" & ratedifs$taxa=="Birds"])` birds and `r round(ratedifs$dif[ratedifs$biome2 == "Tropical forest" & ratedifs$taxa=="Amphibians"])` amphibians in tropical forests and by only `r round(ratedifs$dif[ratedifs$biome2 == "Boreal" & ratedifs$taxa=="Birds"])` birds and `r round(ratedifs$dif[ratedifs$biome2 == "Boreal" & ratedifs$taxa=="Amphibians"])` amphibians in boreal regions. Within the tropics, increasing canopy height generally yields greater gains in richness in forests than in drier woodlands, savannas, and shrublands, though baseline richness in areas with short vegetation are similar (Fig. \@ref(fig:fig1)). For the same increase in canopy height, the increase in bird richness in tropical forests exceeds the increase in bird richness in tropical woodlands, savannas, and shrublands by `r round(ratedifs$dif[ratedifs$biome2 == "Tropical forest" & ratedifs$taxa=="Birds"]) - round(ratedifs$dif[ratedifs$biome2 == "Tropical woodland, savanna, and shrubland" & ratedifs$taxa=="Birds"])` species. The difference in the capacity for tall canopies to generate biodiversity between densely and sparsely vegetated tropical regions is particularly evident in Madagascar where the global model underpredicts vertebrate richness in the humid forests of the east coast and overpredicts richness in the drier woodlands and shrublands of the southwest (Figs. \@ref(fig:fig1)c, S\@ref(fig:resids-taxa)). Thus, increases in canopy height have greater positive impacts on richness in the wet tropics and lower impacts in drier or higher latitude regions, mirroring impacts of mountains [@rahbekHumboldtsEnigmaWhat2019]. Simply put, taller canopies generate biodiversity globally, but are more effective generators of biodiversity in the wet tropics.

[ADD MODEL TABLES TO SUPPLEMENT]

# The mediating effect of verticality

```{r}
source("scripts/03_analyses/02_vertmaps_and_vertRichness_correlations.R")

# mean proportion of arboreal species
parb.wettropics = round(as.numeric(parb.mean[parb.mean$biome=="Tropical & Subtropical Moist Broadleaf Forests", 2]),2)

The vertical layering of species within forests may mediate the capacity for tall forests to support rich assemblages [@oliveiraVerticalStratificationInfluences2019; @macarthurBirdSpeciesDiversity1961; @xingEcologicalPatternsProcesses2023a]. As forests increase in height, additional arboreal species may be added to the assemblage in novel canopy space [@macarthurBirdSpeciesDiversity1961; @rollLinkingVertebrateSpecies2015; @larueDiversityVolumeRelationships2023]. An increase in the number of arboreal species without an equivalent increase in terrestrial or fossorial species increases the verticality of the assemblage, which represents the skew in the vertical distribution of species toward the canopy [@oliveiraVerticalStratificationInfluences2019; @scheffersVerticalArborealityHorizontal2017]. Within wooded ecosystems, we therefore expect regions exhibiting high verticality to support species rich assemblages and steep increases in richness with canopy height [@oliveiraVerticalStratificationInfluences2019]. To quantify the verticality of an assemblage, we use the aforementioned trait databases, assigning verticality scores to each species based on their occurrence in fossorial (0), terrestrial (0.5), and arboreal (1) habitats [@oliveiraVerticalStratificationInfluences2019; @scheffersVerticalArborealityHorizontal2017]. For species occurring in multiple habitats, we averaged scores across habitats. We then calculated assemblage verticality as the standard effect size of the average verticality score across species in an assemblage, using 100 simulations to determine if assemblages are more vertical than expected by random chance given their richness and the evolutionary history of the biogeographic realm [@oliveiraVerticalStratificationInfluences2019] (Supplementary methods).

Verticality peaked in the wet tropics and declined toward the poles, exhibiting strong correlations with richness across taxa (Fig. \@ref(fig:fig2), S\@ref(fig:vertmean-biome)). Tropical and subtropical moist forests contained assemblages with an average of `r parb.wettropics*100`% of species exhibiting arboreal strategies, corresponding to some of the most species rich regions (Fig. S\@(fig:meanprop-biome)). Though temperate and boreal regions contain tall forests, albeit not quite reaching the heights of the tallest tropical canopies, they contain assemblages with much lower verticality, and higher proportions of species exhibiting terrestrial and fossorial strategies (Fig. \@ref(fig:fig2)). This pattern suggests that the capacity for taller canopies to support higher richness declines when canopies cannot support highly vertical species assemblages. Understanding environmental factors that constrain the evolution of diverse arboreal assemblages is therefore key to unraveling the vertical dimension to Humbolt's enigma.

# Vegetation structure and canopy climates impact the capacity for tall canopies to support vertical communities

To examine the environmental factors influencing assemblage verticality, we used spatial linear mixed models to model assemblage verticality for each vertebrate class as a function of canopy height and its interaction with macroclimatic factors that may inhibit or enhance the capacity for canopies to support diverse arboreal assemblages. Macroclimate variables included those that may induce physiological constraints - minimum temperature of the coldest month, maximum temperature of the warmest month, precipitation of the driest month, and precipitation of the warmest month - as well as diurnal temperature range, which may influence the distinctness of vertical climate banding (Supplementary methods). We additionally included vegetation density in the models, which impacts the availability of habitat volume. While foliage height diversity and temperature and precipitation seasonality may also impact vertical niche opportunities and the distinctness of climate banding, these variables where strongly correlated with canopy height, minimum temperature, and dry season precipitation, respectively, and were therefore excluded from the models. The macroclimate variables included in these models represent climate conditions experienced in open habitats rather than the microclimate conditions across vertical forest profiles [@defrenneWeatherStationsLack2016; @bramerAdvancesMonitoringModelling2018]. While using macroclimate data was necessary due to the lack of global vertically resolved microclimate data, its use inhibits our ability to understand how microclimates experienced by forest-dwelling organisms impact vertical assemblage structure and richness patterns [@klingesProximalMicroclimateMoving2024; @zellwegerForestMicroclimateDynamics2020]. We therefore additionally modeled vertical microclimate profiles in six 1° x 1° grid cells within forested regions from tropical to boreal latitudes. To do so, we integrated information on the vertical distribution of vegetation derived from NASA's GEDI spaceborne lidar mission [@dubayahGlobalEcosystemDynamics2020] with ERA5 macroclimate data at a high temporal resolution [@hersbachERA5HourlyData2023] in a process-based microclimate model that resolves within-canopy micrometeorological processes [@macleanMicroclimfFastCanopy2024; @macleanMicropointCanopyPoint2025]. To reduce variability in vertical profiles not attributed to latitudinal differences, we restricted microclimate models to points below 500 m a.s.l and that represented the top 75th to 98th quantiles of canopy height within the modeled region (Supplementary methods). By estimating the climate conditions experienced by forest-dwelling organisms, these models provide insight into the microclimatic processes that underlie macroclimate patterns and allow canopies to support vertically stratified and species rich assemblages.

Our models of verticality indicate that taller canopies support highly vertical vertebrate assemblages, while dry and thermally variable macroclimate conditions, including extreme hot and cold temperatures limit verticality and/or the capacity for taller forests to support assemblages with higher verticality, albeit to varying degrees across vertebrate classes (Fig. \@ref(fig:fig3), Table S\@ref(tab:vert-model-results)). These patterns arise due to impacts of climate on vegetation structure and physiological constraints to arboreal lifestyles, as well as impacts of temporal climate variation on the distinctness of vertical climate bands.

## Climate produces structural constraints to verticality

Climate impacts the structural complexity of vegetation, which plays a key role in the the availability of arboreal niche opportunities and is typically associated with high animal richness [@xingEcologicalPatternsProcesses2023a; @daviesAdvancesAnimalEcology2014]. Our models indicated that high dry season precipitation positively impacts assemblage verticality across taxa (Fig. \@ref(fig:fig3)). Wet climates with low precipitation seasonality enable the development of tall and structurally complex forests with densely layered vegetation and generally high tree diversity [@ehbrechtGlobalPatternsClimatic2021a; @decontoCharacterizingStructuralComplexity2024; @hakkenbergClimateMediatesRelationship2021]. Wet climates also support diverse epiphyte communities, which provide additional microhabitats in the canopy that contribute to high arboreal richness [@taylorVascularEpiphytesContribute2022; @thomsenSecondaryFoundationSpecies2018; @ellwoodDoublingEstimateInvertebrate2004]. Hotspots of structural complexity in wet regions that coincide with some of the tallest forests, including the Choco-Darien, the Amazon, Borneo, and New Guinea [@ehbrechtGlobalPatternsClimatic2021a; @decontoCharacterizingStructuralComplexity2024], thus provide opportunities for fine-scale niche partitioning, yielding species rich assemblages [@camargoForestVerticalComplexity2018; @gouveiaForestStructureDrives2014; @oliveiraVerticalStratificationInfluences2019].
